---
title: "Lab2"
author: "Enter Name"
date: "2023-01-04"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(leaflet)
library(maps)
library(ggmap)
options(scipen=999)
#knitr::opts_chunk$set(echo = FALSE)

```

## Read in the data & get state map setup

```{r dataRead, echo=FALSE, message = FALSE}
# Get state data
stateToGet <- 'CA'
iceCream <- read.csv('https://raw.githubusercontent.com/kwartler/Hult_Visualizing-Analyzing-Data-with-R/main/DD1/B_Mar2/lab/mcBroken_august_12_2022.csv')

# Get the full state name from the input
fullStateName <- state.name[match(stateToGet,state.abb)]

# Get a single state for some sections
oneStateIceCream <- subset(iceCream, iceCream$state==stateToGet)

# Get map data
us            <- fortify(map_data('state'), region = 'region')
oneState      <- us[ us$region %in% tolower(fullStateName), ]
# make sure oneState has rows!  Otherwise subset it manually!

```


## Examine the data 
<details>
<summary>Show Example of the Data</summary>
```{r dataHead, echo=FALSE}
knitr::kable(head(iceCream))
```
</details>

## 1. Plot all locations on a map

<details>
<summary>Data Prep Instructions</summary>
- Not needed this time!
</details>

<details>
<summary>Plot Instructions</summary>  
- Create a `ggplot` map
     - Initialize with `ggplot()`
     - Add a `geom_map` layer where the `data` is `oneState`
          - The `map` parameter should be `oneState`
          - The `aes`thetics have a `map_id` of `region` & `group=group`
          - Outside of the `aes` add a `"white"` `fill` parameter & `"black"` `color`.
          - Lastly, add `linewidth = 0.25` to declare the line size
    - Add the curved coordinate system for maps with `coord_map('albers', lat0 = 39, lat1 = 45)`
    - Declare a theme with `theme_map()` layer
    - Next, add the points layer with `geom_point` referring to `oneStateIceCream`.
         - The aesthetics have x as the `lon`, y as `lat`.  Additionally declare the `color` with the variable `is_broken` & change the `shape` with the same variable as another parameter.
         - Outside of aesthetics, add `0.35` transparency with `alpha` to help with overplotting.
    - Add a title using `paste0` & the previously created variable `stateToGet` with `' Ice Cream Machine Status'` to programmatically update the title.
</details>

<details><summary>Show Plot</summary>

```{r locations, echo=FALSE}
____() + 
  ____(____ = ____, 
       ____ = ____, aes(____ = ____, ____  = ____), 
       ____ = '____', ____ = '____', ____ = ____) + 
  ____(____,____, ____) +
  ____() +
  ____(data  = ____, 
             aes(____, ____, ____, ____), 
             alpha = ____) + 
  ggtitle(paste0(____, ____))

```
</details>

## 2. Zoom map with tile

<details>
<summary>Data Prep Instructions</summary>
- Create `maxCity` with `table` function applied to the `city` column of `oneStateIceCream`.  Nest this inside `as.data.frame` to change the object class.
- Overwrite `maxCity` using `maxCity` with square brackets `[` & `]` where `which.max` is applied to the `Freq` column to the left of the comma.  Here we retain the first column only (right of the comma) 
- Let's `subset` the `oneStateIceCream` data into a new object called `recentDay`.  Here we use two logical conditions.
     - First, let's subset when the `date` column is equal to the `max` date
     - Second, let's subset when the `city` column is equal to the `maxCity` variable.
- In order to map a map using Open Street Maps we need to programmatically declare the bounding box with our data.  This information is stored in a named vector called `areaBoundingBox`.  *Instead of just copying, pasting, pause to conceptually understand the relationship of lat/lon and how you would draw a box that encompasses all the points in the data.*
     - First, is upper left as `left` which is the `min` `lon`gitude of `recentDay`
     - Next, the bottom (left side) is declared as the `max` of `lat`titude
     - Moving to the `right` we switch to the `max` of `lon`
     - Lastly, the `top` is declared as the `max` of `lat`
</details>

<details>
<summary>Plot Instructions</summary>  
- Initialize the `ggmap` while nesting `get_stamenmap` to access OSM.  For the Stamen map pass in the `boundareaBoundingBox` coordinates.  Then declare a `zoom` integer & the optional `maptype` as `"toner-lite"`
- Add the theme `void` 
- Add the `geom_point` layer passing the `data` `recentDay`.  The `aes`thetics are:
     - x-axis is `lon`
     - y-axis is `lat`
     - color the points by the `is_broken` variable
- Add a title layer `paste`ing the `maxCity` variable to `'McDonald Locations'`
</details>

<details><summary>Show Plot</summary>

```{r locationsZoom, echo=FALSE}
# Get the city with the most stores
maxCity <- ____(____(oneStateIceCream$____))
maxCity <- ____[which.max(maxCity$____),1]

# Get the most recent day & only the max city
recentDay <- ____(oneStateIceCream, 
                    oneStateIceCream$____ == max(oneStateIceCream$____)
                    & oneStateIceCream$____ == ____)

# Programmatically declare the bounding box
areaBoundingBox <- c(left   = ____, 
                     bottom = ____, 
                     right  = ____, 
                     top    = ____)

# Create the plot
ggmap(get_stamenmap(____, ____, ____ = "toner-lite")) + 
  ____() +
  geom_point(data=____, aes(____))  + 
  ____(____)
```
</details>

## 3. Thermometer Plot

*Could also be done as a pct stacked bar but this is another view for learning*

<details>
<summary>Data Prep Instructions</summary>
- Create `breaks` using `aggregate` declaring `is_broken` by `city` along with `oneStateIceCream` & using the function `sum`
- Create `storeObservations` by nesting `table` inside `as.data.frame` referring to `oneStateIceCream`'s `city` (if you forgot the syntax see previous section)
     - This is then nested *again* within `setNames` where the first column is renamed to `"city"` for the next step.
- Create the `plotDF` by joining both `breaks` and `storeObservations` using `inner_join`.  Since the joining column was declared previously as `city` it can be left blank but is left here to demonstrate how to declare the column names explicitly (and its a best practice)  
- Since the data is quite large, the `plotDF` is re-declared using `order` to obtain the 10 cities with the most number of broken machines.
</details>

<details>
<summary>Plot Instructions</summary>  
- Create a `ggplot` with `plotDF` as the `data`.
- Add the base thermometer with `geom_col`. The `aes` are
     - `x` needs to have `reorder`ed the `city` column by `-Freq`. 
     - `y` needs to have `Freq`
     - Outside of aesthetics, `fill` the bars as `grey` & `width` as `0.5`
- Add the red thermometers with another `geom_col`
     - The `x` is the same as the other columns so they align.
     - Now `y` is declared as `is_broken`
     - Outside of the `aes` call, `fill` is `#f00b42` & `width` is `0.15`
- Add a `theme` that rotates the `axis.text.x` `90` degrees.  Hint, see the first lab, stackoverflow or chatGPT 
- Add a title as `"City Ice Cream Machines Uptime Vs Broken"`
</details>

<details><summary>Show Plot</summary>
```{r thermometer, echo=FALSE}
# Sum the broken times across cities within the period of the data
breaks <- ____(____~____, ____, ____)

# Count the observations in the period by the city
____ <- ____(as.data.frame(table(oneStateIceCream$city)), c('____','____')) 

# Join for the plot
plotDF <- ____(____,storeObservations, by = c('____'='city'))

# Take the top 10 by broken (cities with most McDonald's broken machines)
plotDF <- plotDF[____(plotDF$is_broken, ____)[1:10],]

# Create the plot
ggplot(____) + 
  ____()+
  geom_col(aes(____), ____) + 
  geom_col(aes(____),____), ____) +
  ____(____) + 
  ggtitle(____)
```
</details>

## 4. Leaflet Interactive Map for all locations

<details>
<summary>Data Prep Instructions</summary>
- Declare the `pal`ette object using `colorFactor` across the factor `iceCream$is_broken`.  The colors should be `darkgreen` & `darkred`.
</details>

<details>
<summary>Plot Instructions</summary>  
- Declare a `leaflet` object with the `iceCream` data.
- Add the tiles with a blank layer `addTiles`
- Add the location circles with `addCircles`
     - Adjust the mouse-over with the `popup` parameter.  This will use limited HTML in a `paste` call.  The `<br>` is an HTML break which renders neatly in the map webpage. Examine the names of the `iceCream` data frame to identify the appropriate columns to use in the popups.
     - After the `popup` parameter, add `color` declaring the `~pal` custom function applied to the `is_broken` colum.
</details>

<details><summary>Show Plot</summary>
```{r leafletMap, echo=FALSE}
# Define the marker colors
pal <- ____(palette = c('____', '____'),
                   domain = ____)

# Create the interactive map
leaflet(data = ____) %>%
  addTiles() %>% 
   addCircles(popup = paste("Location:", iceCream$____, "<br>",
                            "City:", iceCream$____,"<br>",
                            "Date:",iceCream$____ ,"<br>",
                            "Broken:", iceCream$____),
              color = ~pal(____)) 

# You can also add clusters & change to markers but 
# custom colors for clusters does not appear to be supported.  
# Try out this type of last layer instead:
#addMarkers( popup = paste("Location:", ____, "<br>",
#                            "City:", ____,"<br>",
#                            "Date:",____ ,"<br>",
#                            "Broken:", ____),
#              clusterOptions = markerClusterOptions())
```

## 5. Timelines

<details>
<summary>Data Prep Instructions</summary>
- Create `worstCities` using `aggregate` with `is_broken` by `city` & `date`.  Apply this formula to `iceCream` & `sum`
- The data is `order`ed by `is_broken` within square brackets along with `decreasing=T` & the second worst city is pre-selected.  You can change this to examine other cities.  This new object is called `badCity` 
- Create `plotDF` by creating a Boolean vector with the `%in%` operator.  Check whether the `iceCream` `city` vector contains the `city` name identified in `badCity`
</details>

<details>
<summary>Plot Instructions</summary>  
- Initialize the `ggplot` with `plotDF`
- Add `geom_point` layer
     - Use `last_checked` as the x-axis
     - Declare `y` as the unique `street` variable.
     - Adjust the `shape` to the `is_broken` class.
     - Adjust the `color` to the `is_broken` class.
- Adjust the `theme` so that the `axis.text.x` `size` is `1.5`
- Add a `minimal` `theme`
- Add a title using `paste` & `'Sensor Reads by Time & Location at:',plotDF$city[1]` to programmatically adjust the title.
     
</details>

<details><summary>Show Plot</summary>
```{r timeLine, echo=FALSE}
# Sum the broken reads by city and date to make a more interesting subset
worstCities <- ____(____)
badCity <- worstCities[____(worstCities$is_broken, ____)[2],] 

# Subset the original data to the city being investigated
plotDF <- ____[____$city %____% ____$____,] 

ggplot(data = ____) +
  geom_point( aes(x= last_checked, y = street, shape = is_broken, color = is_broken))+
  theme(____=element_text(____)) +
  ____() +
  ggtitle(____)
```

## 6. Heatmap

<details>
<summary>Data Prep Instructions</summary>
- `onlyUS` can be created using the membership [`%in%`] operator.  Check if the `iceCream$state` value has membership in the built in `state.abb` object.  This will drop non-US state locations from our data set as a cleaning step.
- To aid our learning with a more compelling heatmap let's create more date observations.  This is left intact in the code but is explained here:
<br>
<mark>"The `min` is applied to the `onlyUS$date` data column to identify the oldest date in the data.  This is nested in `seq` with `by` `"day"` & `length.out` to get a full `7` days from the originating date.  Finally, `sample` is applied to the `7` dates along with the `size = nrow(onlyUS)` integer.  This programmaticaly declares the number needed rather than explicit in case the data object changes later.  The last parameter `replace = T` is needed since the number of days has length `7` which is shorter than the number of rows needed.  All this is needed to create a date vector called `fakeDate`.  This is left-hand appended as a column in the `onlyUS` data as `$fakeDate`")
</mark>
<br>
- Create the `plotDF` using `aggregate` to `sum` the `onlyUS` `is_broken` data by `state` & the synthetic column `fakeDate`
</details>

<details>
<summary>Plot Instructions</summary>  
- Initialize `ggplot` with the `plotDF` data.
- Add the `geom_tile` layer with:
    - `x` as `fakeDate`
    - `y` as `state`
    - `fill` as `is_broken`
- Add the `bw` theme
- To ensure all dates are shown in the x-axis add a layer `scale_x_date()`.  Within this layer,  declare the format as `date_labels="%d %b %y"`.  *This follows the format of the `strptime` from base-r.  Here %d is day, %b is the month abbreviation name and %y is 2 digit year.  Explore these by changing `b` to `m` or `y` to `Y`.  If you type `?strptime` into console you will see the official documentation.* Add another parameter here as `breaks = unique(onlyUS$fakeDate)` to declare the unique x-axis sections.
- The y-axis has WY at the top so this needs to be revered.  One way to do this is by adding `scale_y_discrete` as a layer with `limits=rev` to reverse the order.
- Add a `theme` layer so that the x and y axis titles are blank.  Again use stackoverflow, chatgpt or previous examples for help.
- Add a layer `scale_fill_viridis_c()` to adjust the colors to a color-blind acceptable palette. 
- Add a title `"Fake Ice cream Outages by Date"`

</details>

<details><summary>Show Plot</summary>
```{r heatmap, echo=FALSE}
# Subset using a prebuilt R object 
onlyUS <- iceCream[____$state %in% ____,]

# Since this is a toy data set we will add some randomness to dates to improve learning
fakeDate <- sample(seq(min(onlyUS$date), by = "day", length.out = 7), 
                   size    = nrow(onlyUS), 
                   replace = T)
onlyUS$fakeDate <- as.Date(fakeDate)

# Get the plot data summing broken by state and the fake dates
plotDF <- ____(____)

# Build the heatmap
ggplot(data = ____) + 
  geom_tile(aes(x = fakeDate, y = state, fill= is_broken)) + 
  ____() +
  scale_x_date(date_labels=____,breaks = ____(onlyUS$____)) +
  scale_y_discrete(____) +
  theme(____, ____) +  
  ____()+
  ggtitle(____)

```
